---
title: "Subset Calculations"
output:
  html_document:
    theme: united
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
  word_document:
    toc: yes
    toc_depth: '2'
---

# Overview
This document will demonstrate how to perform calculations with subsets of data. <br>
For example, calculating the means for both male and female samples.

# Outline
1. Using the `aggregate()` function to compute means in subsets of one or two group levels.
2. Using the Tidy syntax to compute means in subsets of one or two group levels.

# Dataset
We will use these data for our examples:
```{r}
carsdata <- mtcars

set.seed(123) # this function with the specific number "123" will make sure
              # we generate the same random numbers
scoredata <- data.frame(
  gender = rep(c("Male", "Female"), times = 1, each = 1),
  score = round(rnorm(120, 70, 5)),
  school = rep(c(1,2,2,1), times = 1, each = 1)
)
```
For the cars dataset, we will calculate the mean "mpg" by the "carb" (number of carburetors) group levels.
```{r}
head(carsdata)
```
For the score dataset, we will calculate the mean score for each gender variable.
```{r}
head(scoredata)
```

# Using `aggregate()`
## Subset by One Group levels
The `aggregate()` function is simple function that lets you use a single function for a variable by the group level of other categorical variables.
```{r}
# template:
# aggregate(what you want to calculate with ~ what you want to subset by, dataset, function)
aggregate(mpg ~ vs, carsdata, mean)
```
In this case, we calculated the mean MPG by the levels of the carb variable.

Next, is the same function used to calculate the mean score for each gender in the scores dataset.
```{r}
# template:
# aggregate(what you want to calculate with ~ what you want to subset by, dataset, function)
aggregate(score ~ gender, scoredata, mean)
```
Keep in mind, the `aggregate()` function does not only calculate means. You can calculate other things like median, and standard deviations by specifying it in the function arguments. (i.e., replace mean with median)

Note: The aggregate function generates a dataframe. You can assign a name to that dataframe and call it like your other dataframes

```{r}
# specifying what data you want from the dataframe created by the aggregate function
aggregate(score ~ gender, scoredata, mean)$score[1] # notice the $score[1] usage
meanScores <- aggregate(score ~ gender, scoredata, mean)
meanScores
meanScores$score[1]
```

## Subset by 2 Group levels
You can subset using two group variables. <br>
Using the cars data, we will calculate the mean MPG by each "carb" levels and each "vs" levels.
```{r}
aggregate(mpg ~ vs*gear, carsdata, mean)
```

Using the scores dataset, we will calculate the mean score for each gender levels and school levels.
```{r}
aggregate(score ~ gender*school, scoredata, mean)
```

# Using Tidy Method

```{r, message=F}
# import the tidyverse Package
library(tidyverse)
```

## Subset by One Group levels
Using the Tidyverse package is good option if you are familiar with the Tidy syntax. <br>
The `aggregat()` function can feel limited because it can only run one function (i.e., mean or median). <br>
However, using the tidy codes, you can write one chain of code to calculate aggregates of various kinds.
```{r}
scoredata %>%           # score data frame at the start of the pipe
  group_by(gender) %>%  # specify which variable to subset (group) by
  summarise(meanScore = mean(score),     # specify column names and what you want
            medianScore = median(score), # those columns to contain
            sdScore = sd(score))         # i.e., mean, median, sd...
```
```{r}
carsdata %>%
  group_by(vs) %>%
  summarise(meanMPG = mean(mpg),
            medianMPB = median(mpg),
            sdMPG = sd(mpg))
```
Like with the `aggregate()` function, these codes generate a dataframe that you can assign a name to, and you can interact with them like any other dataframe.
```{r}
SummScore1 <- scoredata %>% 
  group_by(gender) %>% 
  summarise(meanScore = mean(score),
            medianScore = median(score),
            sdScore = sd(score))

# Calling the mean score value for males
SummScore1$meanScore[SummScore1$gender == "Male"]
```

## Subset by Two Group levels
Simply, add more group variables inside the `group_by()` function when you want to subset with more group levels.
```{r}
scoredata %>%
  group_by(gender, school) %>%  # When you want to subset by more than one 
                                # group levels, just add those group variables
                                # inside the group_by function
  summarise(meanScore = mean(score), 
            medianScore = median(score),
            sdScore = sd(score))
```

```{r}
carsdata %>%
  group_by(vs, gear) %>%
  summarise(meanMPG = mean(mpg),
            medianMPB = median(mpg),
            sdMPG = sd(mpg))
```
Again, these codes generate a dataframe that you can assign a name to, and you can interact with them like any other dataframe.
```{r}
SummScore2 <- scoredata %>% # Assigning a name to the summary dataframe
  group_by(gender, school) %>%
  summarise(meanScore = mean(score), 
            medianScore = median(score),
            sdScore = sd(score))

# Calling the mean score value for females in school 2
SummScore2$meanScore[SummScore2$gender == "Female" & SummScore2$school == 2]
```