---
title: "Calculating with Columns for Each Row"
output: 
  html_document:
    theme: united
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r, echo=F, message=F}
library(dplyr)
library(kableExtra)
```

# Overview
Some people were having trouble performing calculations with a series of columns for each row. This document will demonstrate how this can be done.

# Outline
1. Calculating the row sum of a series of columns.
2. Calculating the row mean of a series of columns.
3. Using the `apply()` function

For this demonstration, we will be looking at the mtcars data set (only the first 8 rows) which is already available to you.

```{r}
cars <- head(mtcars, 8)
```

```{r, echo=F}
kbl(cars) %>%
  kable_styling(full_width = F) %>%
  column_spec(9:12, background = "#a8dfff")
```

# Using the `rowSums()` Function
We will use the `rowSums()` function to calculate the sum of values in a series of columns in our cars data frame. <br>
Specifically, we will create a new column called "sums" which will contain the sum of the last 4 columns (vs, am, gear, carb) for each row.

```{r}
cars$sums <- rowSums(cars[8:11], na.rm = T) # "na.rm" is optional
```

```{r, echo=F}
kbl(cars) %>%
  kable_styling(full_width = F) %>%
  column_spec(13, background = "#f5ff51") %>%
  column_spec(9:12, background = "#a8dfff")
```
<br>

### Finding the Column Index Number
It is vary likely that you have a data frame with many more columns and you do not know the column index number.

In this case, you can use the following code to find the index number of a specific column.

```{r}
which(colnames(cars) == "vs")
```
<br>
With these codes, you could construct do something like the following. <br>
We are merely assigning the column numbers to variables that we name "start" and "end". <br>
This will produce an equivalent result as the previous code. <br>
However, we do not need to know the index number of the columns we want to sum up.

```{r}
# The following codes do the same thing.

# option 1:
start <- which(colnames(cars) == "vs")
end <- which(colnames(cars) == "carb")

cars$sums <- rowSums(cars[start:end])

# option 2:
cars$sums <- rowSums(
  cars[which(colnames(cars) == "vs"):which(colnames(cars) == "carb")]
)
```
Note: It might not be vary efficient to use this method. You might save time just running the `which(colnames(cars) == "vs")` code and keeping the numbers in mind instead of making new variables for them.


```{r, echo=F}
kbl(cars) %>%
  kable_styling(full_width = F) %>%
  column_spec(13, background = "#f5ff51") %>%
  column_spec(9:12, background = "#a8dfff")
```

# Using the `rowMeans()` Function
The procedure for calculating the means of a series of column values is almost identical with the previous section. <br> <br>

We will create a new column named "means", and it will contain the mean of the same columns as before.

```{r}
# The following codes do the same thing.

# option 1:
cars$means <- rowMeans(cars[8:11], na.rm = T)

# option 2:
start <- which(colnames(cars) == "vs")
end <- which(colnames(cars) == "carb")

cars$means <- rowMeans(cars[start:end])

# option 3:
cars$means <- rowMeans(
  cars[which(colnames(cars) == "vs"):which(colnames(cars) == "carb")]
)
```

```{r, echo=F}
kbl(cars) %>%
  kable_styling(full_width = F) %>%
  column_spec(14, background = "#f5ff51") %>%
  column_spec(9:12, background = "#a8dfff")
```

# Using the `apply()` function
In the previous sections, I introduced you to the `rowSums()` and the `rowMeans()` functions.
However, these functions are only limited to finding the sums and means respectively.
<br> <br>

There will be times when you want to perform a similar calculations with a different function (e.g., median, standard deviation, maximum/minimum values, etc...)
<br> <br>

To do this we will use the `apply()` function. We will first replicate the previous sections and move of to other functions.

## Sum
```{r}
# The following codes do the same thing.

# option 1:
# template: apply(data, MARGIN = 1 for row / 2 for col, function, 
#                 optional arguments of the function being used ("sum", in this case))
cars$sums2 <- apply(cars[8:11], MARGIN = 1, sum, na.rm = T)
                                        # note that the parentheses are absent
                                        # after the "sum" function
# MARGIN = 1 indicates that the sum function will be run on each row

# option 2:
start <- which(colnames(cars) == "vs")
end <- which(colnames(cars) == "carb")

cars$sums2 <- apply(cars[start:end], MARGIN = 1, sum, na.rm = T)

# option 3:
cars$sums2 <- apply(
  cars[which(colnames(cars) == "vs"):which(colnames(cars) == "carb")],
  MARGIN = 1, sum, na.rm = T)

```

## Mean
```{r}
# The following codes do the same thing.

# option 1:
# template: apply(data, MARGIN = 1 for row / 2 for col, function,
#                 optional arguments of the function being used (mean, in this case))
cars$means2 <- apply(cars[8:11], MARGIN = 1, mean, na.rm = T)
                                        # note that the parentheses are absent
                                        # after the "mean" function
# MARGIN = 1 indicates that the sum function will be run on each row

# option 2:
start <- which(colnames(cars) == "vs")
end <- which(colnames(cars) == "carb")

cars$means2 <- apply(cars[start:end], MARGIN = 1, mean, na.rm = T)

# option 3:
cars$means2 <- apply(
  cars[which(colnames(cars) == "vs"):which(colnames(cars) == "carb")],
  MARGIN = 1, mean, na.rm = T)

```

```{r, echo=F}
kbl(cars) %>%
  kable_styling(full_width = F) %>%
  column_spec(c(15,16), background = "#f5ff51") %>%
  column_spec(9:12, background = "#a8dfff") %>%
  remove_column(2:8)
```

## Other Functions
```{r}
# median
cars$median <- apply(cars[8:11], MARGIN = 1, median, na.rm = T)

# standard deviation
cars$sd <- apply(cars[8:11], MARGIN = 1, sd, na.rm = T)

# maximum value
cars$max <- apply(cars[8:11], MARGIN = 1, max, na.rm = T)

# minimum value
cars$min <- apply(cars[8:11], MARGIN = 1, min, na.rm = T)
```

```{r, echo=F}
kbl(cars) %>%
  kable_styling(full_width = F) %>%
  column_spec(17:20, background = "#f5ff51") %>%
  column_spec(9:12, background = "#a8dfff") %>%
  remove_column(2:8)
```

# Conclusion
For calculating the sums and means of a series of column for each row, I recommend using the `rowSums()` and `rowMeans()` functions since they were specifically designed to do these calculations. <br> 

For other calculations, you can use the `apply()` function with the "MARGIN = 1" argument and by specifying the appropriate function to do the calculation.

## Alternative:
The R package: matrixStats exists and it provides more functions like the `rowSums()` function. <br>

Check out their website and reference manual if you are interested.
<br>
Webpage: https://github.com/HenrikBengtsson/matrixStats
<br>
Reference: https://cran.r-project.org/web/packages/matrixStats/matrixStats.pdf



